Errata

### 2025-04-06 tex/Manuscript-5.tex
Fix typos and update references in Methodology section of the manuscript. Corrected "Köhn1929" to "Kohn1929" and updated citation for the $bestNormalize$ package from Peterson2021 to Peterson2020.

A critical aspect of the plant-associated contribution to these soil processes occurs at the root-soil interface, i.e., within the rhizosphere. As plants grow, their roots physically alter soil minerals, and root-associated microorganisms release weathering agents (e.g., organic acids; \citealp{Uroz2009}) that mobilize essential soil nutrients \citep{Gregory2022}. 
In addition, the release of rhizodeposits at the root-soil interface represents a major pathway for the build-up of rather persistent mineral-associated OM (MAOM) in soil \citep{Sokol2019, Villarino2021}. Moreover, the increased abundance of labile C in the rhizosphere stimulates the growth and activity of microorganisms. This in turn drives the formation of aggregates via microbially derived binding agents (i.e., extracellular polymeric substances; EPS) or via the enmeshment facilitated by fungal hyphae \citep{Chenu2011, Costa2018}. In addition, root exudates (e.g., root mucilage or polysaccharide exudates) can either adhere directly to mineral surfaces or function as glue-like binding agents, binding soil particles together, forming MAOM \citep{Roetzer2023} and driving soil aggregation \citep{Baumert2018}. As such, roots substantially shape the chemical, physical, and biological nature of the soil at the root-soil interface, and the rhizosphere is widely recognized as a dynamic system in both space and time \citep{Jones2009, Hinsinger2009, Kuzyakov2019}. However, relatively little is known about the temporal changes in processes occurring in the rhizosphere, as well as potential long-term effects of the living root system that may persist beyond the lifetime of the roots \citep{Oliver2021}. These effects, so-called 'root legacy effects', might have implications for soil physical, chemical, and biological parameters that can remain in the soil after plant death and thus shape the subsequent detritusphere \citep{Wurst2015}.

@@

Here, we examine the interactions of root OM inputs with soil aggregate formation and the fate of SOM into POM and MAOM at the root-soil interface in living and decaying root systems of a semiarid top- and subsoil. In order to simulate these processes during the early stages of soil formation, two incubation phases were conducted, mimicking the natural growth process of a pioneer perennial herbaceous plant species (\textit{Helenium aromaticum} (Hook.) L. H. Bailey); the first phase with living plants ('rhizosphere phase') with rhizodeposits as the main C input, followed by a decomposition phase ('detritusphere phase'), with plant and root litter as the main C input source (\cref{fig:supp_fig_s1_label}--\cref{fig:supp_fig_s2_label}). While numerous studies have contributed to our understanding of biogeochemical and microbial processes in both the rhizosphere and detritusphere (e.g., \citealp{Poll2008}; \citealp{Marschner2012}; \citealp{Tian2013}; \citealp{Vidal2021}), these systems are often studied separately. This means that the succession from a living to a decaying root system is not included, nor are the potential legacy effects from the rhizosphere that could influence processes in the developing detritusphere \citep{Wurst2015}. Furthermore, potential interactions between rhizodeposits and root litter as the two main root C inputs to SOM are unknown. To address this research gap, we employed an experimental design in which the transition from rhizosphere to root detritusphere, as well as the establishment of the root-associated microbiome, remained undisturbed. Replicates were sampled at two time points, at the end of the rhizosphere and detritusphere phase, after which the distribution of water-stable aggregates and their contribution to the total C and N pool were quantified by wet-sieving. The fate and chemical composition (C, N, \textsuperscript{13}C and \textsuperscript{15}N) of OM in POM (free POM (fPOM), occluded POM (oPOM and oPOM<\SI{20}{\micro\metre})) and MAOM was determined by density fractionation and \textsuperscript{13}C CP-MAS NMR. Lignin-derived phenols were extracted from plant and root biomass, and the structure of root-associated microbial communities in the rhizosphere and detritusphere was characterized by phospholipid fatty acid (PLFA) extractions. We hypothesized that roots would promote the formation of water-stable macroaggregates in both topsoil and subsoil, and that root-induced aggregation would remain after plant death in the detritusphere as root legacy effects. We assumed that microbial abundance would increase in the rhizosphere, particularly in the C-poor subsoil, due to the increased abundance of labile C around the roots. We further hypothesized that the transition to the detritusphere, as labile C compounds are depleted and replaced by more complex litter-derived compounds, would induce a succession in the root-associated microbiome, specifically towards increasing Gram-positive bacterial abundance.

@@

Soil was sampled from five subplots on south-facing top slopes within the Santa Gracia Natural Reserve, located in the Coquimbo Region in Chile (\ang{29;44;26.3}S, \ang{71;09;01.2}W) with mean annual precipitation of \SI{89}{\milli\metre\,\year^{-1}} and mean annual air temperature of \SI{13.7}{\degreeCelsius} \citep{MinisterioObrasPublicas2017}. All five subplots were classified as Cambisol according to the WRB \citep{IUSSWorkingGroupWRB2015}. Vegetation was dominated by shrubs (\SIrange{30}{40}{\percent} coverage; including \textit{Proustia cuneifolia}, \textit{Balbisia peduncularis}, \textit{Cordia decandra} and \textit{Baccharis paniculatum}), classified as 'Interior Mediterranean desert scrub' \citep{Luebert2006}, and heavily disturbed by grazing animals \citep{Oeser2018}. The sampling site was selected as a suitable representation of an initial dryland soil with low vegetation cover, low organic C content, shallow and weakly structured A horizon \citep{Bernhard2018}. Topsoil and subsoil material was collected and sieved (\SI{< 2}{\milli\metre}) from the profile walls of A and B horizons (\SIrange{0}{3}{\centi\metre} and \SIrange{5}{30}{\centi\metre}, respectively), representing soils at different stages of development (\cref{tab:supp_table_s1_label}). This allowed us to directly compare topsoil and subsoil, but also to use the subsoil as a model system for initial soil formation via plant-soil interactions in a saprolitic material.

@@

The topsoil and subsoil material from the five subplots was homogenized and cleared of plant residues and seeds to prevent unwanted germination during incubation. The two soil materials (topsoil and subsoil) were weighed separately into pots (\SI{350}{\centi\metre\cubed}) aiming to resemble bulk density according to field conditions (\SIrange{1.6}{1.7}{\gram\per\centi\metre\cubed}; \citealp{Bernhard2018}). Pre-germinated seeds (\textit{Helenium aromaticum} (Hook.) L. H. Bailey) were sown on half of the pots, while the other half was left free of plants and hereafter referred to as controls (\cref{fig:supp_fig_s1_label}). After the initial incubation period of 70 days under greenhouse conditions ('rhizosphere phase'), half of the samples were harvested, while the remaining half were placed in complete darkness and incubated at room temperature for 100 days ('detritusphere phase'). The samples were left completely undisturbed during decomposition, meaning that above-ground biomass was left to decompose on the soil surface. During both incubation phases, the samples received \SI{6}{\milli\liter} of rainwater per day, without correcting for potential differences in soil moisture, e.g. due to differences in plant water use between treatments.

After incubation, the plants were cut just above the soil surface and the roots were separated from the bulk soil (\cref{fig:supp_fig_s2_label}). To minimize artifact effects of biocrust on the soil surface, the top \SIrange{0}{1}{\centi\metre} of all samples was discarded \citep{Weber2022}. The soil sticking to the roots was collected separately and hereafter referred to as 'root-adhering soil'. Since the pots were thoroughly rooted, the remaining soil in the pots was defined as 'rhizosphere soil'. During the detritusphere phase, there was not enough soil adhering to the decaying roots to sample separately, meaning that all the soil in the pots were collected as 'detritusphere soil' from this phase. After incubation, replicate samples were combined in pairs (two by two) to form composite replicates. As such, there were initially 12 replicates per treatment during the incubation phases (48 pots in total), which later resulted in 6 composite replicates for each treatment for subsequent analyses. After harvest, plant and root samples were immediately frozen in liquid N. A homogenized aliquot of each soil sample was freeze-dried and stored at \SI{-20}{\degreeCelsius} for later microbial analyses (section 2.3). The remaining bulk soil material was air-dried at room temperature. Plant and root samples were analyzed for C, N, \textsuperscript{13}C, \textsuperscript{15}N (section 2.6), and lignin-derived phenols were extracted (section 2.8). Basic soil parameters (pH and electrical conductivity) were determined for soil material before and after each incubation period, and the gravimetric water content was determined for incubated samples (\cref{tab:supp_table_s1_label}).

@@

We tested all parameters for homoscedasticity using Levene's test and the residuals of each model were tested for Normality using QQ-Plots and the Shapiro-Wilk test. In cases where the assumption of normality or homoscedasticity was not met, the data was appropriately transformed. To evaluate differences between the data presented in \cref{tab:table_1_label}, we conducted a one-way analysis of variance (ANOVA) with Tukey's honestly significant differences (HSD) as a post-hoc test. The remaining data was analyzed using mixed-effect models (package: \texttt{nlme}) using soil material (topsoil vs subsoil) and phase (rhizosphere phase vs detritusphere phase) as fixed factors and the ID as a random effect. All statistical testing was conducted in R (version 4.2.1; \citealp{RCoreTeam2008}) using RStudio (Version 2022.07.2, R Core Team, Vienna, Austria; \citealp{RStudioTeam2015}). Further, data shown in text and tables are given as the mean values and standard deviations.

@@

The soil material clearly impacted the growth of the above- (plant) and belowground (root) biomass, with lower plant and root biomass and lower C and N contents in the subsoil compared to the topsoil (\cref{tab:table_1_label}). The soil material also influenced the elemental composition of belowground biomass, with broader C:N ratios in roots growing in the subsoil (\num{54.54 \pm 3.99}) compared to the topsoil (\num{34.43 \pm 6.69}; \(P < 0.001\)). The difference in root litter quality between the growing materials was further underlined by the higher lignin\textsubscript{NMR}:N ratio (\num{295.67 \pm 14.96} in subsoil compared to \num{207.66 \pm 36.13} in topsoil; \(P= 0.001\)). The decomposition of aboveground biomass was more progressed compared to belowground biomass, with approximately \SI{7}{\percent} of the biomass remaining after the detritusphere phase in both soil materials (compared to \(\sim\)\SI{22}{\percent} and \SI{13}{\percent} of roots remaining in topsoil and subsoil, respectively; \cref{tab:table_1_label}).

The molecular mixing model according to \citet{Nelson2005} showed how the composition of both above- and belowground biomass changed with decomposition, with the concentration of lignin\textsubscript{NMR} increasing from \SI{22.30 \pm 0.96}{\percent} to \SI{37.11 \pm 4.57}{\percent} in decomposing roots in the topsoil (\(P < 0.001\); \cref{tab:table_1_label}). This was accompanied by a slight decrease in carbohydrate\textsubscript{NMR} (from \SI{57.59 \pm 2.31}{\percent} to \SI{44.60 \pm 8.41}{\percent}; \(P= 0.008\)). These patterns differed from the aboveground biomass, where instead the relative protein\textsubscript{NMR} content almost doubled with decomposition (\(P=0.004\)), while the lignin content remained unaffected. The chemical composition of the decaying biomass in the subsoil could not be determined due to insufficient sample material remaining after decomposition.

The increased lignin\textsubscript{NMR} in decaying root biomass in the topsoil was corroborated by a slight, however not significant, increase in lignin monomers (from \SI{121.03 \pm 47.54}{mg\,VSC\,\gram^{-1}\,C} in living roots to \SI{205.80 \pm 81.72}{mg\,VSC\,\gram^{-1}\,C} in decaying roots; \(P=0.24\)). The progressed decomposition of the roots in the topsoil was reflected in an increased (Ac/Al)\textsubscript{V} ratio (from \num{0.10 \pm 0.02} to \num{0.87 \pm 0.23}; \citealp{KoegelKnabner1988}) and overall, in changed lignin composition as indicated by the changing S:V, C:V and Ac:Al ratios. This was not reflected in the aboveground biomass (\cref{tab:table_1_label}).

@@

The introduction of a pioneer plant to the system did not notably change the elemental composition of the soil, neither in the rhizosphere phase nor in the detritusphere phase, with no significant changes in C and N content or in the C:N ratio compared to the controls (\cref{fig:supp_fig_s3_label}; \cref{tab:supp_table_s2_label}). Overall, C and N decreased and C:N ratios narrowed over the course of the incubation with no clear distinction between planted and control samples (\cref{fig:supp_fig_s3_label}).

Stable isotope measurements only showed minor shifts in the natural abundance of \textsuperscript{15}N and \textsuperscript{13}C in root-adhering subsoil, which was slightly enriched in \textsuperscript{15}N (\(P=0.011\)) and depleted in \textsuperscript{13}C (\(P=0.006\)) compared to the bare soil controls (\cref{fig:supp_fig_s4_label}). The decrease in \(\delta\)\textsuperscript{13}C was also weakly reflected in the topsoil (\(P=0.065\)). Overall, the two soil materials differed strongly in their isotopic composition, with the subsoil enriched in \textsuperscript{13}C but depleted in \textsuperscript{15}N compared to the topsoil (\cref{fig:supp_fig_s4_label}).

@@

The living roots fostered the formation of macroaggregates in the topsoil, which was reflected in an increased mass contribution of the \SIrange{250}{2000}{\micro\metre} aggregate size obtained by submerged sieving (\SI{71.18 \pm 3.90}{\percent} compared to \SI{64.38 \pm 1.67}{\percent} in controls, \(P=0.001\); \cref{fig:supp_fig_s5_label}). This was associated with lower mass contributions of the smaller aggregate size classes compared to controls (\(P < 0.001\) for large microaggregates and \(P=0.081\) for small microaggregates). These patterns were only weakly reciprocated in the subsoil. Similarly, while the patterns persisted into the detritusphere incubation phase, the differences were no longer significant.

>>>>

A critical aspect of the plant-associated contribution to these soil processes occurs at the root-soil interface, i.e., within the rhizosphere. As plants grow, their roots physically alter soil minerals, and root-associated microorganisms release weathering agents (e.g., organic acids; \citep{Uroz2009}) that mobilize essential soil nutrients \citep{Gregory2022}. 
In addition, the release of rhizodeposits at the root-soil interface represents a major pathway for the build-up of rather persistent mineral-associated OM (MAOM) in soil \citep{Sokol2019, Villarino2021}. Moreover, the increased abundance of labile C in the rhizosphere stimulates the growth and activity of microorganisms. This in turn drives the formation of aggregates via microbially derived binding agents (i.e., extracellular polymeric substances; EPS) or via the enmeshment facilitated by fungal hyphae \citep{Chenu2011, Costa2018}. In addition, root exudates (e.g., root mucilage or polysaccharide exudates) can either adhere directly to mineral surfaces or function as glue-like binding agents, binding soil particles together, forming MAOM \citep{Roetzer2023} and driving soil aggregation \citep{Baumert2018}. As such, roots substantially shape the chemical, physical, and biological nature of the soil at the root-soil interface, and the rhizosphere is widely recognized as a dynamic system in both space and time \citep{Jones2009, Hinsinger2009, Kuzyakov2019}. However, relatively little is known about the temporal changes in processes occurring in the rhizosphere, as well as potential long-term effects of the living root system that may persist beyond the lifetime of the roots \citep{Oliver2021}. These effects, so-called 'root legacy effects', might have implications for soil physical, chemical, and biological parameters that can remain in the soil after plant death and thus shape the subsequent detritusphere \citep{Wurst2015}.

@@

Here, we examine the interactions of root OM inputs with soil aggregate formation and the fate of SOM into POM and MAOM at the root-soil interface in living and decaying root systems of a semiarid top- and subsoil. In order to simulate these processes during the early stages of soil formation, two incubation phases were conducted, mimicking the natural growth process of a pioneer perennial herbaceous plant species (\textit{Helenium aromaticum} (Hook.) L. H. Bailey); the first phase with living plants ('rhizosphere phase') with rhizodeposits as the main C input, followed by a decomposition phase ('detritusphere phase'), with plant and root litter as the main C input source. While numerous studies have contributed to our understanding of biogeochemical and microbial processes in both the rhizosphere and detritusphere (e.g., \citep{Poll2008}; \citep{Marschner2012}; \citep{Tian2013}; \citep{Vidal2021}), these systems are often studied separately. This means that the succession from a living to a decaying root system is not included, nor are the potential legacy effects from the rhizosphere that could influence processes in the developing detritusphere \citep{Wurst2015}. Furthermore, potential interactions between rhizodeposits and root litter as the two main root C inputs to SOM are unknown. To address this research gap, we employed an experimental design in which the transition from rhizosphere to root detritusphere, as well as the establishment of the root-associated microbiome, remained undisturbed. Replicates were sampled at two time points, at the end of the rhizosphere and detritusphere phase, after which the distribution of water-stable aggregates and their contribution to the total C and N pool were quantified by wet-sieving. The fate and chemical composition (C, N, \textsuperscript{13}C and \textsuperscript{15}N) of OM in POM (free POM (fPOM), occluded POM (oPOM and oPOM<\SI{20}{\micro\metre})) and MAOM was determined by density fractionation and \textsuperscript{13}C CP-MAS NMR. Lignin-derived phenols were extracted from plant and root biomass, and the structure of root-associated microbial communities in the rhizosphere and detritusphere was characterized by phospholipid fatty acid (PLFA) extractions. We hypothesized that roots would promote the formation of water-stable macroaggregates in both topsoil and subsoil, and that root-induced aggregation would remain after plant death in the detritusphere as root legacy effects. We assumed that microbial abundance would increase in the rhizosphere, particularly in the C-poor subsoil, due to the increased abundance of labile C around the roots. We further hypothesized that the transition to the detritusphere, as labile C compounds are depleted and replaced by more complex litter-derived compounds, would induce a succession in the root-associated microbiome, specifically towards increasing Gram-positive bacterial abundance.

@@

Soil was sampled from five subplots on south-facing top slopes within the Santa Gracia Natural Reserve, located in the Coquimbo Region in Chile (\ang{29;44;26.3}S, \ang{71;09;01.2}W) with mean annual precipitation of \SI{89}{\milli\metre\,\year^{-1}} and mean annual air temperature of \SI{13.7}{\degreeCelsius} \citep{MinisterioObrasPublicas2017}. All five subplots were classified as Cambisol according to the WRB \citep{IUSSWorkingGroupWRB2015}. Vegetation was dominated by shrubs (\SIrange{30}{40}{\percent} coverage; including \textit{Proustia cuneifolia}, \textit{Balbisia peduncularis}, \textit{Cordia decandra} and \textit{Baccharis paniculatum}), classified as 'Interior Mediterranean desert scrub' \citep{Luebert2006}, and heavily disturbed by grazing animals \citep{Oeser2018}. The sampling site was selected as a suitable representation of an initial dryland soil with low vegetation cover, low organic C content, shallow and weakly structured A horizon \citep{Bernhard2018}. Topsoil and subsoil material was collected and sieved (\SI{< 2}{\milli\metre}) from the profile walls of A and B horizons (\SIrange{0}{3}{\centi\metre} and \SIrange{5}{30}{\centi\metre}, respectively), representing soils at different stages of development. This allowed us to directly compare topsoil and subsoil, but also to use the subsoil as a model system for initial soil formation via plant-soil interactions in a saprolitic material.

@@

The topsoil and subsoil material from the five subplots was homogenized and cleared of plant residues and seeds to prevent unwanted germination during incubation. The two soil materials (topsoil and subsoil) were weighed separately into pots (\SI{350}{\centi\metre\cubed}) aiming to resemble bulk density according to field conditions (\SIrange{1.6}{1.7}{\gram\per\centi\metre\cubed}; \citep{Bernhard2018}). Pre-germinated seeds (\textit{Helenium aromaticum} (Hook.) L. H. Bailey) were sown on half of the pots, while the other half was left free of plants and hereafter referred to as controls. After the initial incubation period of 70 days under greenhouse conditions ('rhizosphere phase'), half of the samples were harvested, while the remaining half were placed in complete darkness and incubated at room temperature for 100 days ('detritusphere phase'). The samples were left completely undisturbed during decomposition, meaning that above-ground biomass was left to decompose on the soil surface. During both incubation phases, the samples received \SI{6}{\milli\liter} of rainwater per day, without correcting for potential differences in soil moisture, e.g. due to differences in plant water use between treatments.

After incubation, the plants were cut just above the soil surface and the roots were separated from the bulk soil. To minimize artifact effects of biocrust on the soil surface, the top \SIrange{0}{1}{\centi\metre} of all samples was discarded \citep{Weber2022}. The soil sticking to the roots was collected separately and hereafter referred to as 'root-adhering soil'. Since the pots were thoroughly rooted, the remaining soil in the pots was defined as 'rhizosphere soil'. During the detritusphere phase, there was not enough soil adhering to the decaying roots to sample separately, meaning that all the soil in the pots were collected as 'detritusphere soil' from this phase. After incubation, replicate samples were combined in pairs (two by two) to form composite replicates. As such, there were initially 12 replicates per treatment during the incubation phases (48 pots in total), which later resulted in 6 composite replicates for each treatment for subsequent analyses. After harvest, plant and root samples were immediately frozen in liquid N. A homogenized aliquot of each soil sample was freeze-dried and stored at \SI{-20}{\degreeCelsius} for later microbial analyses (section 2.3). The remaining bulk soil material was air-dried at room temperature. Plant and root samples were analyzed for C, N, \textsuperscript{13}C, \textsuperscript{15}N (section 2.6), and lignin-derived phenols were extracted (section 2.8). Basic soil parameters (pH and electrical conductivity) were determined for soil material before and after each incubation period, and the gravimetric water content was determined for incubated samples.

@@

We tested all parameters for homoscedasticity using Levene's test and the residuals of each model were tested for Normality using QQ-Plots and the Shapiro-Wilk test. In cases where the assumption of normality or homoscedasticity was not met, the data was appropriately transformed. To evaluate differences between the data presented in \ref{tab:M5-T1}, we conducted a one-way analysis of variance (ANOVA) with Tukey's honestly significant differences (HSD) as a post-hoc test. The remaining data was analyzed using mixed-effect models (package: \texttt{nlme}) using soil material (topsoil vs subsoil) and phase (rhizosphere phase vs detritusphere phase) as fixed factors and the ID as a random effect. All statistical testing was conducted in R (version 4.2.1; \citep{RCoreTeam2008}) using RStudio (Version 2022.07.2, R Core Team, Vienna, Austria; \citep{RStudioTeam2015}). Further, data shown in text and tables are given as the mean values and standard deviations.

\begin{sidewaystable}[htbp]
  \centering
  \scriptsize
  \caption{Elemental and isotopic composition of above- and belowground biomass. Relative amounts of carbohydrates, proteins, lignin, lipids, and carbonyls from two incubation phases in topsoil and subsoil. Asterisks denote significance ($\bullet$ $p < 0.1$, *$p < 0.05$, **$p < 0.01$, ***$p < 0.001$).}
  \begin{tabular}{llcccccccc}
  \toprule
  \multirow{3}{*}{Category} & \multirow{3}{*}{Property} & \multicolumn{4}{c}{\textbf{Topsoil}} & \multicolumn{4}{c}{\textbf{Subsoil}} \\
  \cmidrule(lr){3-6} \cmidrule(lr){7-10}
  & & \multicolumn{2}{c}{Rhizosphere} & \multicolumn{2}{c}{Detritusphere} & \multicolumn{2}{c}{Rhizosphere} & \multicolumn{2}{c}{Detritusphere} \\
  \cmidrule(lr){3-4} \cmidrule(lr){5-6} \cmidrule(lr){7-8} \cmidrule(lr){9-10}
  & & Above & Below & Above & Below & Above & Below & Above & Below \\
  \midrule
  \multirow{8}{*}{Basic properties}
  & C (mg g$^{-1}$) & 4.06 $\pm$ 0.20 & 3.70 $\pm$ 0.08 & 3.49 $\pm$ 0.47 & 3.97 $\pm$ 0.74 & 3.16 $\pm$ 0.50 & 3.70 $\pm$ 0.09 & 3.23 $\pm$ n.a. & 3.36 $\pm$ 1.43 \\
  & N (mg g$^{-1}$) & 0.17 $\pm$ 0.05** & 0.11 $\pm$ 0.02** & 0.25 $\pm$ 0.06** & 0.11 $\pm$ 0.02** & 0.09 $\pm$ 0.01 & 0.07 $\pm$ 0.00 & 0.11 $\pm$ n.a. & 0.10 $\pm$ 0.04 \\
  & C:N & 25.61 $\pm$ 7.77** & 34.43 $\pm$ 3.43** & 14.67 $\pm$ 3.99*** & 37.26 $\pm$ 8.89 & 33.58 $\pm$ 2.90 & 53.54 $\pm$ 5.27*** & 28.70 $\pm$ n.a. & 32.37 $\pm$ -- \\
  & Biomass (g) & 0.47 $\pm$ 0.17*** & 0.44 $\pm$ 0.13*** & 0.11 $\pm$ 0.07*** & 0.03 $\pm$ 0.02*** & 0.25 $\pm$ 0.09*** & 0.29 $\pm$ 0.07*** & 0.04 $\pm$ n.a. & 0.02 $\pm$ 0.00*** \\
  & Total biomass (g) & 0.90 $\pm$ 0.22*** & 0.13 $\pm$ 0.09*** & 0.54 $\pm$ 0.11*** & 0.05 $\pm$ 0.07*** & \multicolumn{4}{c}{--} \\
  & Root:shoot & 1.07 $\pm$ 0.48** & 0.32 $\pm$ 0.20** & 1.27 $\pm$ 0.37 & 0.24 $\pm$ n.a. & \multicolumn{4}{c}{--} \\
  & C (mg) & 1.91 $\pm$ 0.78*** & 1.62 $\pm$ 0.51*** & 0.38 $\pm$ 0.29*** & 0.12 $\pm$ 0.09*** & 0.75 $\pm$ 0.21 & 1.07 $\pm$ 0.12*** & 0.36 $\pm$ n.a. & 0.06 $\pm$ 0.04*** \\
  & N (mg) & 0.07 $\pm$ 0.02*** & 0.05 $\pm$ 0.01*** & 0.03 $\pm$ 0.02*** & 0.00 $\pm$ 0.01*** & 0.02 $\pm$ 0.01 & 0.02 $\pm$ 0.00 & 0.01 $\pm$ n.a. & 0.00 $\pm$ 0.00 \\
  & $\delta^{15}$N ($\permil$ air N$_2$) & 4.96 $\pm$ 0.78 & 4.14 $\pm$ 0.59$^{\bullet}$ & 4.29 $\pm$ 0.71 & 3.22 $\pm$ 0.81$^{\bullet}$ & 4.66 $\pm$ 0.61 & 4.41 $\pm$ 0.45*** & 3.94 $\pm$ n.a. & 1.56 $\pm$ 0.53*** \\
  & $\delta^{13}$C ($\permil$ V-PDB) & -29.84 $\pm$ 0.78 & -28.68 $\pm$ 0.51 & -29.46 $\pm$ 1.34 & -29.50 $\pm$ 1.01 & -31.94 $\pm$ 0.39 & -30.52 $\pm$ 0.35 & -31.22 $\pm$ n.a. & -31.66 $\pm$ 1.76 \\
  \midrule
  \multicolumn{10}{l}{\textbf{Relative contents from mixing model (\%)}} \\
  & Carbohydrate$_{\text{NMR}}$ & 53.68 $\pm$ 7.07$^{\bullet}$ & 57.69 $\pm$ 2.31** & 42.67 $\pm$ 8.17$^{\bullet}$ & 44.60 $\pm$ 8.41** & 47.38 $\pm$ 4.89 & 63.13 $\pm$ 5.09 & 49.78 $\pm$ n.a. & 37.17 $\pm$ n.a. \\
  & Protein$_{\text{NMR}}$ & 10.76 $\pm$ 3.47** & 7.46 $\pm$ 1.36 & 19.81 $\pm$ 6.16** & 7.39 $\pm$ 2.22 & 7.48 $\pm$ 0.69 & 4.57 $\pm$ 0.35 & 9.36 $\pm$ n.a. & 7.39 $\pm$ n.a. \\
  & Lignin$_{\text{NMR}}$ & 21.05 $\pm$ 2.04 & 22.30 $\pm$ 0.98** & 25.10 $\pm$ 4.79 & 37.11 $\pm$ 4.57*** & 25.52 $\pm$ 2.46 & 20.54 $\pm$ 1.93 & 25.62 $\pm$ n.a. & 42.83 $\pm$ n.a. \\
  & Lipid$_{\text{NMR}}$ & 8.04 $\pm$ 1.28 & 5.93 $\pm$ 0.89 & 9.37 $\pm$ 1.85 & 3.42 $\pm$ 0.89 & 7.97 $\pm$ 0.92 & 4.63 $\pm$ 0.90 & 8.76 $\pm$ n.a. & 1.66 $\pm$ n.a. \\
  & Carbonyl$_{\text{NMR}}$ & 6.46 $\pm$ 2.99 & 6.62 $\pm$ 0.98** & 3.04 $\pm$ 2.41 & 7.49 $\pm$ 3.01** & 11.66 $\pm$ 2.18 & 7.13 $\pm$ 2.16 & 6.48 $\pm$ n.a. & 10.94 $\pm$ n.a. \\
  & Lignin:N & 132.24 $\pm$ 34.21 & 207.66 $\pm$ 36.13$^{\bullet}$ & 106.31 $\pm$ 31.66 & 351.58 $\pm$ 80.547$^{\bullet}$ & 275.88 $\pm$ 41.97 & 295.67 $\pm$ 14.96 & 236.74 $\pm$ n.a. & 428.00 $\pm$ n.a. \\
  \midrule
  \multicolumn{10}{l}{\textbf{Lignin monomers}} \\
  & VSC (mg g$^{-1}$ C) & 58.89 $\pm$ 32.17 & 121.03 $\pm$ 47.54 & 51.96 $\pm$ 20.66 & 205.80 $\pm$ 81.72 & 39.37 $\pm$ 10.21 & 89.12 $\pm$ 40.70 & 40.41 $\pm$ n.a. & n.a. $\pm$ n.a. \\
  & S:V & 0.44 $\pm$ 0.10 & 0.66 $\pm$ 0.12** & 0.42 $\pm$ 0.19 & 1.56 $\pm$ 0.41** & 0.29 $\pm$ 0.06 & 0.77 $\pm$ 0.30 & 0.42 $\pm$ n.a. & n.a. $\pm$ n.a. \\
  & C:V & 0.19 $\pm$ 0.06 & 0.24 $\pm$ 0.03** & 0.24 $\pm$ 0.14 & 0.09 $\pm$ 0.01*** & 0.20 $\pm$ 0.14 & 0.18 $\pm$ 0.12 & 0.26 $\pm$ n.a. & n.a. $\pm$ n.a. \\
  & Ac:Al & 0.34 $\pm$ 0.08 & 0.32 $\pm$ 0.03* & 0.50 $\pm$ 0.23 & 0.55 $\pm$ 0.04* & 0.39 $\pm$ 0.27 & 0.22 $\pm$ 0.15 & 0.48 $\pm$ n.a. & n.a. $\pm$ n.a. \\
  & Ac:Al$_V$ & 0.11 $\pm$ 0.03 & 0.10 $\pm$ 0.02*** & 0.18 $\pm$ 0.04 & 0.87 $\pm$ 0.23*** & 0.11 $\pm$ 0.07 & 0.07 $\pm$ 0.03 & 0.15 $\pm$ n.a. & n.a. $\pm$ n.a. \\
  \bottomrule
  \end{tabular}
  \label{tab:M5-T1}
  \end{sidewaystable}

@@

The soil material clearly impacted the growth of the above- (plant) and belowground (root) biomass, with lower plant and root biomass and lower C and N contents in the subsoil compared to the topsoil (\ref{tab:M5-T1}). The soil material also influenced the elemental composition of belowground biomass, with broader C:N ratios in roots growing in the subsoil (\num{54.54 \pm 3.99}) compared to the topsoil (\num{34.43 \pm 6.69}; \(P < 0.001\)). The difference in root litter quality between the growing materials was further underlined by the higher lignin\textsubscript{NMR}:N ratio (\num{295.67 \pm 14.96} in subsoil compared to \num{207.66 \pm 36.13} in topsoil; \(P= 0.001\)). The decomposition of aboveground biomass was more progressed compared to belowground biomass, with approximately \SI{7}{\percent} of the biomass remaining after the detritusphere phase in both soil materials (compared to \(\sim\)\SI{22}{\percent} and \SI{13}{\percent} of roots remaining in topsoil and subsoil, respectively; \ref{tab:M5-T1}).

The molecular mixing model according to \citet{Nelson2005} showed how the composition of both above- and belowground biomass changed with decomposition, with the concentration of lignin\textsubscript{NMR} increasing from \SI{22.30 \pm 0.96}{\percent} to \SI{37.11 \pm 4.57}{\percent} in decomposing roots in the topsoil (\(P < 0.001\); \ref{tab:M5-T1}). This was accompanied by a slight decrease in carbohydrate\textsubscript{NMR} (from \SI{57.59 \pm 2.31}{\percent} to \SI{44.60 \pm 8.41}{\percent}; \(P= 0.008\)). These patterns differed from the aboveground biomass, where instead the relative protein\textsubscript{NMR} content almost doubled with decomposition (\(P=0.004\)), while the lignin content remained unaffected. The chemical composition of the decaying biomass in the subsoil could not be determined due to insufficient sample material remaining after decomposition.

The increased lignin\textsubscript{NMR} in decaying root biomass in the topsoil was corroborated by a slight, however not significant, increase in lignin monomers (from \SI{121.03 \pm 47.54}{mg\,VSC\,\gram^{-1}\,C} in living roots to \SI{205.80 \pm 81.72}{mg\,VSC\,\gram^{-1}\,C} in decaying roots; \(P=0.24\)). The progressed decomposition of the roots in the topsoil was reflected in an increased (Ac/Al)\textsubscript{V} ratio (from \num{0.10 \pm 0.02} to \num{0.87 \pm 0.23}; \citep{KoegelKnabner1988}) and overall, in changed lignin composition as indicated by the changing S:V, C:V and Ac:Al ratios. This was not reflected in the aboveground biomass (\ref{tab:M5-T1}).

@@

The introduction of a pioneer plant to the system did not notably change the elemental composition of the soil, neither in the rhizosphere phase nor in the detritusphere phase, with no significant changes in C and N content or in the C:N ratio compared to the controls. Overall, C and N decreased and C:N ratios narrowed over the course of the incubation with no clear distinction between planted and control samples.

Stable isotope measurements only showed minor shifts in the natural abundance of \textsuperscript{15}N and \textsuperscript{13}C in root-adhering subsoil, which was slightly enriched in \textsuperscript{15}N (\(P=0.011\)) and depleted in \textsuperscript{13}C (\(P=0.006\)) compared to the bare soil controls. The decrease in \(\delta\)\textsuperscript{13}C was also weakly reflected in the topsoil (\(P=0.065\)). Overall, the two soil materials differed strongly in their isotopic composition, with the subsoil enriched in \textsuperscript{13}C but depleted in \textsuperscript{15}N compared to the topsoil.

@@

The living roots fostered the formation of macroaggregates in the topsoil, which was reflected in an increased mass contribution of the \SIrange{250}{2000}{\micro\metre} aggregate size obtained by submerged sieving (\SI{71.18 \pm 3.90}{\percent} compared to \SI{64.38 \pm 1.67}{\percent} in controls, \(P=0.001\)). This was associated with lower mass contributions of the smaller aggregate size classes compared to controls (\(P < 0.001\) for large microaggregates and \(P=0.081\) for small microaggregates). These patterns were only weakly reciprocated in the subsoil. Similarly, while the patterns persisted into the detritusphere incubation phase, the differences were no longer significant.

Root-induced aggregate formation was further emphasized by elevated organic C and N contents within the larger macroaggregates in the rhizosphere, e.g., from \num{5.09} to \SI{5.80}{mg\,C\,\gram^{-1}}\,fraction in topsoil (\(P=0.026\)) and from \num{1.35} to \SI{1.76}{mg\,C\,\gram^{-1}}\,fraction in subsoil (\(P=0.014\); \ref{fig:M5-F1}a and \ref{fig:M5-F1}b). The C and N contribution of the fractions per g soil further emphasized the distinction of the larger macroaggregates, with higher contents of C and N (\si{mg\,\gram^{-1}}\,soil) in both topsoil and subsoil (\ref{fig:M5-F1}c and \ref{fig:M5-F1}d).
